True Stellar Obliquity Distribution of Exoplanets

### 2022-11-19 src/tex/ms.tex
[showyourwork] overleaf sync

% Abstract with filler text
\begin{abstract}
We build a hierarchical Bayesian framework to constrain true stellar obliquity ($\psi$) distribution in exoplanetary systems. Using the sky-projected stellar obliquities ($\lambda$) of a sample of 162 exoplanets, we infer their $\cos{\psi}$ distribution using a four-parameter Beta mixture model with the assumption of a uniform cosine of stellar inclination ($i_\star$) distribution for each planet. We find $\cos{\psi}$ has a significant pile-up near 1, suggesting the majority of exoplanetary systems are well aligned. For misaligned systems, $\cos{\psi}$ is nearly uniformly distributed between $-0.75$ to $0.75$ with no significant pipe-up near $\cos{\psi} = 0$, suggesting the stellar spin axis is randomly distributed around the planet's orbital axis.
Our Bayesian framework could reproduce the excess of perpendicular planets observed in systems with constrained $i_\star$. However, we find the sample of misaligned systems is limited by small number statistics and likely biased because misaligned systems are often found in hot-star systems, whereas the gravity-darkening method, a technique commonly used to constrain the inclinations of hot stars ($T_{\rm eff} > 7000$K), is only sensitive to detect perpendicular planets. 
\end{abstract}

@@

Stellar obliquity $\psi$ is described as the angle between a planet's orbital axis $\hat{i}_{\rm orb}$ and its host star's spin axis $\hat{i}_{\star}$.
If both $\hat{i}_{\rm orb}$ and $\hat{i}_{\star}$ are independent, random unit vectors with uniform distribution on a three-dimensional sphere, $\cos{\psi}$ will be uniformly distributed between $-1$ and $1$.
In reality, the sky-projected stellar obliquity $\lambda$ can only be measured in transiting exoplanet systems, limiting $\hat{i}_{\rm orb}$ close to $\sim 90\degr$.

Stellar obliquity $\psi$ can be calculated from $\hat{i}_{\rm orb}$ and $\hat{i}_{\star}$ as the following:
\begin{equation}
    \cos{\psi} = \sin{i_\star}\sin{i_{\rm orb}}\cos{\lambda} + \cos{i_\star}\cos{i_{\rm orb}},
\end{equation}
where $i_{\rm orb}$ and $i_\star$ are the projections of $\hat{i}_{\rm orb}$ and $\hat{i}_{\star}$ along the observer's line of sight, and $\lambda$ is the sky-projected stellar obliquity. When $i_{\rm orb} = 90\degr$, $\cos{\psi} = \sin{i_\star}\cos{\lambda}$.

In most exoplanetary systems, only the sky-projected obliquity $\lambda$ is constrained. We could then infer $\psi$ given the measured $\lambda$ and assuming a randomly distributed $i_\star$ (uniformly distributed in $\cos{i_\star}$).
In some systems, $i_\star$ can be constrained via photometric rotational modulation, gravitational darkening or asteroseismology. In such cases, an informative $i_\star$ prior can be used. See Figure\ref{hbm} as an illustration.

Using a sample of stars with $i_\star$, Albrecht+21 found the majority of planets are aligned. For misaligned planets, their stellar obliquities are preferentially distributed between 80--125$\degr$. Furthermore, they found the $\cos{\psi}$ distribution of misaligned systems is not uniformly distributed but clustered near 0.

In this work, we build a Bayesian framework to evaluate the distribution of $\cos{\psi}$ and extend the analysis to the full sample exoplanets that have sky-projected stellar obliquity measurements.

\section{Stellar Obliquity Distribution} \label{sec:hbm}
We model the whole population stellar obliquity distribution and also the sub-population with $i_\star$ constrained. See Figure\ref{hbm} as an illustration.

\begin{figure}[ht!]
    \script{hbm.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/hbm.pdf}
        \caption{Hierarchical Bayesian framework to infer stellar obliquity distribution described by hyperparameters $\theta$ from individual system's stellar obliquity $\psi_n$. $\psi_n$ is calculated by the system's sky-projected stellar obliquity $\lambda_n$ and stellar inclination $i_{\star, n}$.}
        \label{fig:hbm}
    \end{centering}
\end{figure}

@@

We find the $i_\star$ sample, there is a pile-up of planets near $-0.25$ whereas, for all planet sample, the distribution is mostly flat. See Figure 2.

@@

% \begin{figure*}[ht!]
%     \script{dist.py}
%     \begin{centering}
%         \includegraphics[width=\linewidth]{figures/dist.pdf}
%         \caption{$\cos{\psi}$ distributions for the sample with $i_\star$ measured and all planets.}
%         \label{fig:dist}
%     \end{centering}
% \end{figure*}

>>>>

% Abstract with filler text
\begin{abstract}
We build a hierarchical Bayesian framework to constrain true stellar obliquity ($\psi$) distribution in exoplanetary systems.
Using the sky-projected stellar obliquities ($\lambda$) of a sample of 151 exoplanets, we infer their $\cos{\psi}$ distribution using a four-parameter Beta mixture model with the assumption of a uniform cosine of stellar inclination ($i_\star$) distribution for each planet. We find $\cos{\psi}$ has a significant pile-up near 1, suggesting the majority of exoplanetary systems are well aligned. For misaligned systems, $\cos{\psi}$ is nearly uniformly distributed between $-0.75$ to $0.75$ with no significant pipe-up near $\cos{\psi} = 0$, suggesting the stellar spin axis is randomly distributed around the planet's orbital axis.
Our Bayesian framework could reproduce the excess of perpendicular planets observed in systems with constrained $i_\star$. However, we find the sample of misaligned systems could be biased because of the small sample size and that misaligned systems are often found around hot stars, whereas the gravity-darkening method, a technique commonly used to constrain the inclinations of hot stars ($T_{\rm eff} > 7000$K), is only sensitive to detect perpendicular planets. 
\end{abstract}

We build a hierarchical Bayesian framework to constrain true stellar obliquity ($\psi$) distribution in exoplanetary systems.

@@

Stellar obliquity $\psi$ is described as the angle between a planet's orbital axis $\bf{i}_{\rm orb}$ and its host star's spin axis $\bf{i}_{\star}$.
If both $\bf{i}_{\rm orb}$ and $\bf{i}_{\star}$ are independent, random unit vectors with uniform distribution on a three-dimensional sphere, $\cos{\psi}$ will be uniformly distributed between $-1$ and $1$.
In reality, the sky-projected stellar obliquity $\lambda$ can only be measured in transiting exoplanet systems, limiting $\bf{i}_{\rm orb}$ close to $\sim 90\degr$.

Stellar obliquity $\psi$ can be calculated from $\bf{i}_{\rm orb}$ and $\bf{i}_{\star}$ as the following:
\begin{equation}
    \cos{\psi} = \sin{i_\star}\sin{i_{\rm orb}}\cos{\lambda} + \cos{i_\star}\cos{i_{\rm orb}},
\end{equation}
where $i_{\rm orb}$ and $i_\star$ are the projections of $\bf{i}_{\rm orb}$ and $\bf{i}_{\star}$ along the observer's line of sight, and $\lambda$ is the sky-projected stellar obliquity. A full derivation can be found in, e.g., \cite{Fabrycky09} Equation (1)--(10).
When setting $i_{\rm orb} = 90\degr$, $\cos{\psi} = \sin{i_\star}\cos{\lambda}$.

In most exoplanetary systems, only the sky-projected obliquity $\lambda$ is constrained. We could still infer $\psi$ given the measured $\lambda$ and assuming a randomly distributed $i_\star$ that is uniformly distributed in $\cos{i_\star}$.
In some systems, $i_\star$ can be constrained via photometric rotational modulation, gravity darkening, or asteroseismology. In such systems, an informative $i_\star$ prior can be used.

\cite{Albrecht21} inferred the true stellar obliquities of a sample of stars that have both stellar inclinations and sky-projected stellar obliquity constraints. They found while the majority of planets have a small $\psi$, misaligned planets have $\psi$ clustered between 80--125$\degr$. Furthermore, they found the $\cos{\psi}$ distribution of misaligned systems is not uniformly distributed but clustered near 0, suggesting the stellar spin axis is not randomly distributed around the planetary orbital axis.

In this work, we build a hierarchical Bayesian framework to evaluate the distribution of $\cos{\psi}$ using the sample in \cite{Albrecht21} and extend the analysis to the full sample of exoplanetary systems that have sky-projected stellar obliquity measurements extracted from the review paper of \cite{Albrecht22}.

\section{Stellar Obliquity Distribution} \label{sec:hbm}
We model the whole population stellar obliquity distribution and also the sub-population with $i_\star$ constrained. See Figure\ref{fig:hbm} as an illustration.

\begin{figure}[ht!]
    \script{hbm.py}
    \includegraphics[width=\linewidth]{figures/hbm.pdf}
    \caption{Hierarchical Bayesian framework to infer stellar obliquity distribution described by hyperparameters $\theta$ from individual system's stellar obliquity $\psi_n$. $\psi_n$ is calculated by the system's sky-projected stellar obliquity $\lambda_n$ and stellar inclination $i_{\star, n}$.}
    \label{fig:hbm}
\end{figure}

@@

We find the $i_\star$ sample, there is a pile-up of planets near $-0.25$ whereas, for the all-planet sample, the distribution is mostly flat. See Figure~\ref{fig:psi_dist}.

@@

\cite{Nagpal22} emphasize the importance of picking the right hyperpriors.

\begin{figure*}[ht!]
    \script{psi_dist.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/psi_dist.pdf}
        \caption{$\cos{\psi}$ distributions.}
        \label{fig:psi_dist}
    \end{centering}
\end{figure*}
